Amicable Contempt

### 2023-12-22 manuscript/appendix.qmd
Add theory illustration

---
title: 'Online appendix for "Amicable Contempt: A Conceptual Framework for Understanding International NGO Behavior in the Era of Closing Civic Space"'
short-title: 'Online appendix for "Amicable Contempt"'
author:
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
bibliography: references.bib
crossref: 
  tbl-title: Appendix Table
  fig-title: Appendix Figure
---

>>>>

---
title: 'Online appendix for "Amicable Contempt: A Conceptual Framework for Understanding International NGO Behavior in the Era of Closing Civic Space"'
short-title: 'Online appendix for "Amicable Contempt"'
author:
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
bibliography: references.json
crossref: 
  tbl-title: Appendix Table
  fig-title: Appendix Figure
---

### 2023-12-22 manuscript/manuscript.qmd
Add theory illustration

---
title: "Amicable Contempt"
subtitle: "A Conceptual Framework for Understanding International NGO Behavior in the Era of Closing Civic Space"
short-title: "Amicable Contempt"
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/truthful-tuna>"
abstract: |
  I use evidence from a global survey of international NGOs to define each INGO-related element of the theory of amicable contempt. I find that INGOs are primarily motivated by their core vision and values, but that they have to balance the pursuit of their missions with instrumental concerns such as fundraising, time, staffing, and collaboration. These concerns both limit and enable INGO activities—without substantial instrumental resources and programmatic flexibility, organizations are unable to carry out their mission, while too much emphasis on resource concerns distracts organizations from their core programming and reduces their effectiveness.

  Amid this internal struggle between mission and money, I then introduce how the civil society regulatory environment of INGO host countries is designed to target each element of this struggle. Gatekeeping regulations limit which organizations are allowed to enter the country and constrain the flexibility of organizations already present in the country. These regulations are primarily defensive and are designed to protect the regime against the potentially destabilizing effects of INGO advocacy and service provision. Program capture regulations, on the other hand, are more offensive and target organizational missions instead of their flexibility. These laws are designed to co-opt and reshape INGO programming to fall more in line with government preferences, allowing regimes to take advantage of INGO activities.

bibliography: references.bib
csl: csl/apa.csl
link-citations: true
word-count: "9,878"
---

@@

```{r libraries-data}
library(tidyverse)
library(scales)
library(tidybayes)
library(marginaleffects)
library(patchwork)
library(glue)
library(janitor)
library(gt)
library(targets)

@@

@@

The space for civil society does indeed appear to be closing, but many INGOs have found creative ways to work around legal restrictions to continue their work. For example, the Cairo Institute for Human Rights Studies (CIHRS) has worked in Egypt since 1993 as a think tank, research center, and hub for human rights advocacy for the larger Middle East. Between 2012–2014, however, the legal environment for INGOs working Egypt became increasingly restricted and CIHRS faced significant threats and legal pressure to end its research. In response, in December 2014 CIHRS moved its regional and international operations to Tunisia.[^1] Civic space had closed in Egypt, but CIHRS found a way around the restrictions and continues to engage in research throughout the region.

@@

An INGO's mission and menu of programs will often conflict with the preferences of the regime. Accordingly, an organization must decide how to adapt its programming to fit the requirements and needs of its host country. This decision can be difficult, pitting conflicting organizational values against each other. If the organization is too antagonistic, contentious, or principled, it runs the risk of expulsion, potentially leaving the INGO unable to continue its work in the country. However if the organization is too cooperative, its selection of programs may not be credible or effective as the INGO kowtows to the dictator's demands.

I use an adaptation of negotiation theory [@Raiffa:1982; @Mayer:2010] to conceptualize this dynamic relationship between governments and INGOs.[^3] I model the interaction between regimes and INGOs in any given issue area (human rights, development, advocacy, etc.) in simple two-dimensional policy space (see @fig:theory-full). Each actor has a preference for a certain level of advocacy or service provision, marked as solid circles in the figure (Regime~ideal~ and INGO~ideal~). Governments set *de jure* policies to prevent INGO programming from going beyond the regime's ideal point, while INGOs will engage in programming that, when possible, fits within the bounds set by the regime. Each actor has a range of flexibility in the implementation of their regulations or their programming, shown in the horizontal lines flanking the ideal points. The space where the two ranges overlap determines both the regime's *de facto* level of restrictions and the INGO's actual mix of programming in that country and constitutes the practical operating environment for the INGO.

![Programming preferences in two-dimensional policy space](../../Graphics/theory-full){#fig:theory-full}

@@

Second, the size of the two actors' overlapping negotiation spaces determines the practical operating environment for INGOs. This space falls along a continuum of restrictiveness, ranging from *tolerance*, or very few legal or actual restrictions on INGO programs, to *expulsion*, or a complete ban of all INGO activities. The larger the practical operating environment, the more tolerant the regime will be about international NGOs working in their regimes, and vice versa—when there is no overlap, an INGO will face expulsion.

![Possible regulatory environments and regime–INGO relationships](../../Graphics/theory-grid){#fig:theory-grid}

The relationship between INGOs and regimes is dynamic, and shifts in INGO flexibility and de facto regime civil society regulations can create different practical operating environments, demonstrated in @fig:theory-grid. For instance, there are two circumstances where INGOs will feel less restricted. First, as seen in the top left quadrant of @fig:theory-grid, if the distance between the two parties' ideal points is short and the negotiation spaces overlap—as with INGOs providing non-contentious programming like disaster relief, humanitarian services, or education—INGOs will be cooperative and face a tolerant operating environment. Because of preference alignment, these INGOs will not need to rely on their flexibility to maintain access to their host countries.

More contentious organizations with preferences that diverge from the regime's ideal point, on the other hand, can also face a tolerant environment as long as they have a large amount of programmatic flexibility (see bottom left quadrant of @fig:theory-grid). For example, @Heiss:2019 explores how human rights INGOs advocate for democratization, political reform, and increased respect for human rights by relying on networks of field offices, funded by dozens of private foundations and government grants—resources that give these organizations the flexibility to remain in their host countries and maintain a fairly tolerant operating environment.

Likewise, there are two circumstances where INGOs will feel restricted, or even expelled. Most often, organizations that face restricted operating environments engage in contentious programming and work in countries where regimes have limited de facto regulatory environments, leaving little room for negotiation or overlap (see bottom right quadrant of @fig:theory-grid). In contrast, INGOs that are naturally less contentious and more prone to compliance can face expulsion if they lack flexibility or it the civil society regulatory environment is too constrained (see top right quadrant of @fig:theory-grid).

@@

Studying INGO responses to legal regulations—particularly for organizations working in highly restrictive countries—poses a difficult methodological challenge. There is extensive cross-country research on the relationship between restrictive governments and *domestic* institutions like civil society [@FrantzKendall-Taylor:2014; @Kendall-TaylorFrantz:2014a; @FrantzEzrow:2011], as well as detailed case studies of specific regimes [@Chen:2012; @Teets:2014; @Hildebrandt:2013], and cross-national quantitative work [@DeMattee; @that-one-group-in-europe], but few studies have examined the motivations and strategies of INGOs themselves.

@@

Complete technical details of the survey are provided in the online appendix. I define an INGO as any organization that is not incorporated as a governmental entity or as a profit-seeking private firm that conducts advocacy, provides services, or otherwise works in at least *one* country other than the one it is headquartered in. Of the 17,590 organizations that potentially received an e-mail invitation to participate in the online survey, 537 completed the survey and 104 provided a partial (but usable) response. This yielded a participation rate of 3.8%, which is admittedly low, but comparable to other internet-based organizational surveys [@BaruchHoltom:2008; @HeissKelley:2016]. Respondents were allowed to answer country-specific questions for multiple countries, thus yielding 641 organization-level responses and 659 country-level responses. The survey included 36 open-ended questions, and most respondents answered between 8–16, providing a wealth of qualitative insights into INGO experiences with their host governments. To give voice to these organizations, I quote extensively from these responses with minimal editing or correction. I guaranteed respondents anonymity to encourage candor in responses and to protect privacy. Accordingly, I assign each respondent a random four-digit ID number, and I strip all potentially identifying information from excerpts and quotes from their responses. I also anonymize any country-specific information, replacing references to country names with general geographic regions.

@@

Respondents span the globe and work on a wide variety of issues and policies. The sample of INGOs have their headquarters spread across 93 different countries (see @fig:hq-countries), with roughly two thirds based in either Europe or the Americas, primarily in the US, UK, and Switzerland (see left panel of @fig:work-hq-regions).

![Countries where INGOs are headquartered](../../Output/figures/3-hq-countries){#fig:hq-countries}

![Regions where INGOs are headquartered vs. regions about which INGOs answered questions](../../Output/figures/3-work-hq-regions){#fig:work-hq-regions}

While comparatively few INGOs are headquartered outside the Global North, these organizations have a worldwide reach. In total, INGOs answered questions about their work in 148 different countries. The regional distribution of INGO target countries reflects their broader scope and is almost a mirror image of these organizations' home regions—nearly two-thirds of respondents answered questions about their work in Africa and Asia (see right panel of @fig:work-hq-regions), the most underrepresented regions for NGO headquarters. This disparity could reflect the fact that INGOs from the Global South without a clear English-language internet presence were excluded from the sample, but the pattern also corresponds to similar findings in other surveys of global NGOs [@Limoncelli:2016; @HeissKelley:2016; @FootToftCesare:2015], reinforcing the idea that Northern-based NGOs tend to focus their work in the South. Organizations also operate in a variety of political contexts. Unsurprisingly, given the concentration of INGOs in the Global North, the vast majority of respondents (90%) are headquartered in democracies. The countries these organizations work in, though, show more diversity in regime type—a third of respondents answered questions about their work in autocratic countries.

@@

Building on these critiques, a new trend emerged that treated transnational actors as simultaneously normative and strategic [@Cooley:2002; @Keck:1999; @Kelley:2009a]. @MitchellSchmitz:2014 summarized this debate with their theory of nonprofit principled instrumentalism, or the notion that INGOs must balance their normative mission-driven principles against the instrumental needs of organizational survival. Unlike governments that work toward producing social value (such as maximizing national security or promoting general welfare) and unlike firms that focus on financial performance (such as maximizing profits for shareholders), nonprofit organizations face a tradeoff between their social mission and their need for resources and must pursue both concurrently—they must "instrumentally pursue their principled objectives within the economic constraints and political opportunity structures imposed by their external environments" [@MitchellSchmitz:2014, 489].

INGOs are motivated primarily by their foundational mission and values, and they structure their strategies around this vision to ensure that their programming furthers their mission [@Moore:2000]. Survey respondents confirm this motivation, describing their core mission both in aspirational terms, including "fighting poverty and holding leaders accountable,"[^4] encouraging "awareness and pro-action towards sustainable development,"[^5] and promoting "democracy and open society" and "governments' respect for the human rights of children";[^6] and through more specific statements, such as "improv[ing] healthcare, specifically neurological services, in the least developed nations,"[^7] or ending "the use of immigration detention around the world."[^8]

Most respondents report a tight link between programs and missions, stating that "[o]ur mission, values and vision are reflected in everything that we do," that "all our work is geared towards the accomplishment of this objective," and that "our approach matches our mission precisely."[^9] For many organizations, the "mission and values are the reason we work on these objectives"[^10]—the normative goals of the organization provide strong motivation to engage in pro-social advocacy and services. Specific programming strategies are also closely linked to INGO missions. For instance, an organization working in South Asia has ambitions to "[b]uild the capacity of local organisations to equip adolescent girls to access their rights and address gender inequality through sport." As expected, much of their programming is centered on sports education and leadership training, but importantly, the organization makes a point that nothing they do is implemented directly: because their core mission is to build local capacity, they ensure that "all the work we do is through local organisations."[^11] Organizations thus work to ensure that their strategies align with and reflect their core missions, visions, and values.

@@

The distance between the government's preferences for a given issue and an INGO's core mission and vision for that issue determine how contentious their relationship will be, ranging from cooperative to antagonistic (see @fig:theory-dot-distance). This distance, in turn, influences how governments regulate INGO work. The alignment of organizational missions and strategies with government preferences, therefore, is crucial for understanding government regulation of INGOs. Organizations working on issues that align with regime preferences will be treated more favorably and with less regulatory pressure than INGOs working on more threatening, less aligned issues.

![Distance between regime and INGO ideal points: contentiousness of relationship](../../Graphics/theory-dot-distance){#fig:theory-dot-distance}

@@

According to @MitchellSchmitz:2014, nonprofit organizations are primarily *driven* by their missions and *constrained* by instrumental concerns: worries about money and other resources get in the way of pursuing deeper normative principles. I argue, however, that this view of instrumentality is insufficient. Instrumental concerns, including time, money, or political connections, are not simply constraints. These resources are what allow INGOs to pursue their principled ideals in the first place—they provide organizations with the flexibility to engage in programming and adapt to the political and regulatory environments they face. The region surrounding the organization's ideal point represents its reserve of *programmatic* flexibility (see @fig:theory-ngo-space), or the organization's ability and willingness to make changes to its programs in response to regulations.

![Space around INGO ideal point: instrumental flexibility](../../Graphics/theory-ngo-space){#fig:theory-ngo-space}

@@

Given their frequency, the regulatory changes INGOs observe do not necessarily correlate with formal statutory developments. These laws give bureaucrats and politicians significant interpretive leeway, however, and allow them to reshape the regulatory environment for international NGOs [@TODO-chaudhry-heiss-2022]. Because of this dynamic discretionary de facto environment, INGOs actively seek out information about changes in laws, rules, and regulations. International NGOs are most likely to hear about new legal changes from other NGOs in the country, both from their local partners and from other foreign organizations. They also receive news about changes directly from government officials, sometimes at inopportune moments. One INGO working in East Africa, unaware of any changes in regulations, discovered new funding requirements for foreign NGOs when requesting permits to expand their operations in the country—the organization ultimately decided to not expand as a result.[^52] INGOs also find out about changes in laws through the media and through internet research, including websites that aggregate information about civil society regulations like Civicus and the International Center for Not-for-Profit Law (ICNL).

@@

[^1]:	Cairo Institute for Human Rights Studies, "After 20 years: CIHRS moves its regional and international programs outside Egypt," accessed August 2, 2017, [http://www.cihrs.org/?p=10298&lang=en](http://www.cihrs.org/?p=10298&lang=en).

@@

[^3]:	The best way to understand the theory is to visit an explorable explanation at [ingorestrictions.org/theory/](https://ingorestrictions.org/theory/) and see the different outcomes dynamically [@Gourarie:2016].

[^4]:	Response 1193.

[^5]:	Response 1267.

[^6]:	Responses 1328 and 1252.

[^7]:	Response 1257.

[^8]:	Response 1376.

[^9]:	Response 1245, 1267, and 1272.

[^10]:	Response 1643.

[^11]:	Response 1799.

@@

>>>>

---
title: "Amicable Contempt"
subtitle: "A Conceptual Framework for Understanding International NGO Behavior in the Era of Closing Civic Space"
short-title: "Amicable Contempt"
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/truthful-tuna>"
abstract: |
  Stuff here

bibliography: references.json
csl: csl/apa.csl
link-citations: true
word-count: "9,878"
---

@@

```{r libraries-data}
library(tidyverse)
library(scales)
library(tidybayes)
library(marginaleffects)
library(sf)
library(ggtext)
library(ggalluvial)
library(patchwork)
library(glue)
library(janitor)
library(countrycode)
library(gt)
library(targets)

@@

tar_load(c(
  survey_orgs, survey_countries, survey_all,
  world_map
))

@@

The space for civil society does indeed appear to be closing, but many INGOs have found creative ways to work around legal restrictions to continue their work. For example, since 1993 the Cairo Institute for Human Rights Studies (CIHRS) worked in Egypt as a think tank, research center, and hub for human rights advocacy for the larger Middle East. Between 2012–2014, however, the legal environment for INGOs working Egypt became increasingly restricted and CIHRS faced significant threats and legal pressure to end its research. In response, in December 2014 CIHRS moved its regional and international operations to Tunisia [@CIHRS:2014]. Civic space had closed in Egypt, but CIHRS found a way around the restrictions and continues to engage in research throughout the region.

@@

An INGO's mission and menu of programs will often conflict with the preferences of the regime. Accordingly, an organization must decide how to adapt its programming to fit the requirements and needs of its host country. This decision can be difficult, pitting conflicting organizational values against each other. If the organization is too antagonistic, contentious, or principled, it runs the risk of expulsion, potentially leaving the INGO unable to continue its work in the country. However if the organization is too cooperative, its selection of programs may not be credible or effective as the INGO bends to the government's demands.

I use an adaptation of negotiation theory [@Raiffa:1982; @Mayer:2010] to conceptualize this dynamic relationship between governments and INGOs. I represent the interaction between regimes and INGOs in any given issue area (human rights, development, advocacy, etc.) in simple two-dimensional policy space (see @fig-theory-full). Each actor has a preference for a certain level of advocacy or service provision, marked as solid circles in the figure. Governments set *de jure* policies to prevent INGO activities from going beyond the regime's ideal point, while INGOs will engage in programming that, when possible, fits within the bounds set by the regime. Each actor has a range of flexibility in the implementation of their regulations or their programming, shown in the horizontal lines flanking the ideal points. The space where the two ranges overlap determines both the regime's *de facto* level of restrictions and the INGO's actual mix of programming in that country and constitutes the practical operating environment for the INGO.

```{r}
#| label: fig-theory-full
#| fig-cap: "Programming preferences in two-dimensional policy space"
#| out-width: "100%"
#| include: true

knitr::include_graphics("graphics-manual/out/theory-full.png", auto_pdf = TRUE)
```

@@

Second, the size of the two actors' overlapping negotiation spaces determines the practical operating environment for INGOs. This space falls along a continuum of restrictiveness, ranging from *tolerance*, or very few legal or actual restrictions on INGO programs, to *expulsion*, or a complete ban of all INGO activities. The larger the practical operating environment, the more tolerant the regime will be about international NGOs working in their regimes, and vice versa—when there is no overlap, an INGO will no longer be allowed to operate in the country.

```{r}
#| label: tbl-theory-grid
#| include: true
#| tbl-cap: "Possible combinations of INGO/regime relationships and practical 
#|   operating environments"

tibble(
  col1 = c(
    "**Cooperative**", rep("", 2),
    "**Antagonistic**", rep("", 2)
  ),
  Tolerance = c(
    "The INGO's core mission is close to regime preferences and the INGO/regime relationship is cooperative. De jure laws may be stricter than the INGO's ideal preferences, but because there is overlap between de facto regulations and INGO programs, the regime tolerates the INGO's work and the organization operates comfortably in the country.",
    "cooperation-tolerance",
    "*Example*: The Red Cross or Oxfam in most countries",
    "The INGO's core mission is far from regime preferences and the INGO/regime relationship is antagonistic. Because there are substantial associational rights and freedom for advocacy, and because the INGO has a range of possible programming, the regime tolerates the INGO's work and the organization operates comfortably in the country.",
    "antagonism-tolerance",
    "*Example*: Amnesty International in the United States"
  ),
  Expulsion = c(
    "The INGO's core mission is close to regime preferences and the INGO/regime relationship should theoreticaly be cooperative. However, because the de facto civil society regulatory environment is highly constricted, and because the INGO has limited programmatic flexibility, the regime severely limits or expels the organization.",
    "cooperation-expulsion",
    "*Example*: Ford Foundation or Mercy Corps in India after 2015",
    "The INGO's core mission is far from regime preferences and the INGO/regime relationship is antagonistic. Because the de facto civil society regulatory environment is highly constricted, the regime restricts, bans, or expels the organization even if the INGO has substantial programmatic flexibility.",
    "antagonism-expulsion",
    "*Example*: Amnesty International with North Korea"
  )
) |>
  group_by(group = "Nature of relationship") |> 
  gt(id = "theory-table") |>
  tab_spanner(
    label = "Practical operating environment",
    columns = c(Tolerance, Expulsion)
  ) |> 
  cols_label(col1 = "") |>
  fmt_markdown(columns = everything(), rows = c(1, 3, 4, 6)) |>
  text_transform(
    cells_body(columns = c(Tolerance, Expulsion), rows = c(2, 5)),
    fn = \(x) local_image(filename = glue("graphics-manual/out/{x}.png"))
  ) |>
  tab_style(
    style = cell_borders(sides = "all", color = NULL),
    locations = cells_body()
  ) |>
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "#D3D3D3",
      weight = px(2)
    ),
    locations = cells_body(columns = everything(), rows = c(4))
  ) |>
  tab_style(
    style = cell_text(v_align = "top"),
    locations = cells_body()
  ) |>
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())
  ) |> 
  tab_style(
    style = cell_fill("#FCF2E6"),
    locations = cells_row_groups()
  ) |> 
  tab_style(
    style = cell_fill("#EFECF4"),
    locations = cells_column_spanners()
  ) |> 
  tab_options(
    quarto.disable_processing = TRUE,
    column_labels.border.top.color = "#ffffff",
    column_labels.font.weight = "bold",
    stub_row_group.border.color = "#ffffff",
    stub_row_group.font.weight = "bold",
    heading.align = "center",
    row_group.as_column = TRUE
  )
```

The relationship between INGOs and regimes is dynamic, and shifts in INGO flexibility and de facto regime civil society regulations can create different practical operating environments, demonstrated in @tbl-theory-grid. For instance, there are two circumstances where INGOs will feel less restricted. First, as seen in the top left quadrant of @tbl-theory-grid, if the distance between the two parties' ideal points is short and the negotiation spaces overlap—as with INGOs providing non-contentious programming like disaster relief, humanitarian services, or education—INGOs will be cooperative and face a tolerant operating environment. Because of preference alignment, these INGOs will not need to rely on their flexibility to maintain access to their host countries.

More contentious organizations with preferences that diverge from the regime's ideal point, on the other hand, can also face a tolerant environment as long as they have a large amount of programmatic flexibility (see bottom left quadrant of @tbl-theory-grid). For example, @Heiss:2019a explores how human rights INGOs advocate for democratization, political reform, and increased respect for human rights by relying on networks of field offices, funded by dozens of private foundations and government grants—resources that give these organizations the flexibility to remain in their host countries and maintain a fairly tolerant operating environment.

Likewise, there are two circumstances where INGOs will feel restricted, or even expelled. Most often, organizations that face restricted operating environments engage in contentious programming and work in countries where regimes have limited de facto regulatory environments, leaving little room for negotiation or overlap (see bottom right quadrant of @tbl-theory-grid). In contrast, INGOs that are naturally less contentious and more prone to compliance can face expulsion if they lack flexibility or it the civil society regulatory environment is too constrained (see top right quadrant of @tbl-theory-grid). For instance, TODO in 2015, less contentious INGOs like Mercy Corps were restricted after India's de facto environment constricted [@Carothers:2015]. No formal laws were changed, but the amount of leeway granted to INGOs shrunk considerably and Modi's government froze assets and cancelled registrations of hundreds of INGOs.

@@

Growing body of work on the government side of the the model, looking at how and why regimes regulate INGOs and why they differentially enforce regulations. Less work has been done on the INGO side… TODO
 There is extensive cross-country research on the relationship between restrictive governments and *domestic* institutions like civil society [@FrantzKendall-Taylor:2014; @Kendall-TaylorFrantz:2014a; @FrantzEzrow:2011], as well as detailed case studies of specific regimes [@Chen:2012; @Teets:2014; @Hildebrandt:2013], and cross-national quantitative work [@DeMattee; @that-one-group-in-europe], but few studies have examined the motivations and strategies of INGOs themselves.

Studying INGO responses to legal regulations—particularly for organizations working in highly restrictive countries—poses a difficult methodological challenge.

@@

Complete technical details of the survey are provided in the appendix. I define an INGO as any organization that is not incorporated as a governmental entity or as a profit-seeking private firm that conducts advocacy, provides services, or otherwise works in at least *one* country other than the one it is headquartered in. Of the 17,590 organizations that potentially received an e-mail invitation to participate in the online survey, 537 completed the survey and 104 provided a partial (but usable) response. This yielded a participation rate of 3.8%, which is admittedly low, but comparable to other internet-based organizational surveys [@BaruchHoltom:2008; @HeissKelley:2016]. Respondents were allowed to answer country-specific questions for multiple countries, thus yielding 641 organization-level responses and 659 country-level responses. The survey included 36 open-ended questions, and most respondents answered between 8–16, providing a wealth of qualitative insights into INGO experiences with their host governments. To give voice to these organizations, I quote extensively from these responses with minimal editing or correction. I guaranteed respondents anonymity to encourage candor in responses and to protect privacy. Accordingly, I assign each respondent a random four-digit ID number, and I strip all potentially identifying information from excerpts and quotes from their responses. I also anonymize any country-specific information, replacing references to country names with general geographic regions.

@@

Respondents span the globe and work on a wide variety of issues and policies. The sample of INGOs have their headquarters spread across 93 different countries (see @fig-hq-countries), with roughly two thirds based in either Europe or the Americas, primarily in the US, UK, and Switzerland (see left panel of @fig-work-hq-regions).

```{r}
#| label: fig-hq-countries
#| fig-width: 6
#| fig-asp: 0.542
#| fig-cap: "Countries where INGOs are headquartered"
#| include: true
# Through lots of trial and error, I found that the best size for this map that
# eliminates extra whitespace on the top/bottom and left/right is 6x3.25, or an
# aspect ratio of 0.542ish

df_hq_countries <- survey_orgs |>
  group_by(Q2.2_iso3) |>
  summarize(num_ngos = n()) |>
  ungroup() |>
  mutate(
    num_ngos_ceiling = ifelse(num_ngos >= 30, 30, num_ngos),
    ngo_presence = ifelse(num_ngos >= 1, "At least one NGO respondent", NA),
    prop = num_ngos / sum(num_ngos, na.rm = TRUE)
  ) |>
  mutate(country_name = countrycode(Q2.2_iso3, "iso3c", "country.name")) |>
  arrange(desc(num_ngos))

map_hq <- world_map |> 
  left_join(df_hq_countries, by = join_by(ISO_A3 == Q2.2_iso3)) |> 
  # Remove Fiji for the sake of plotting
  filter(ISO_A3 != "FJI")

ggplot() +
  geom_sf(
    data = map_hq, aes(fill = num_ngos_ceiling, text = ISO_A3),
    linewidth = 0.1, color = "grey30"
  ) +
  scale_fill_binned(
    low = clrs$Peach[1], high = clrs$Peach[7], na.value = "grey95",
    labels = c(1, 10, 20, "30+"), breaks = c(1, 10, 20, 30),
    name = "Number of NGOs: ",
    guide = guide_colorsteps(
      barwidth = 7, barheight = 0.4,
      title.vjust = 1
    )
  ) +
  coord_sf(crs = "+proj=robin") +
  theme_ingo_map()
```

```{r}
#| label: fig-work-hq-regions
#| include: true
#| fig-width: 4
#| fig-height: 2
#| out-width: "80%"
#| fig-cap: "Regions where INGOs are headquartered vs. regions about which 
#|   INGOs answered questions"

continents_marked <- survey_orgs |>
  select(home = Q2.2_iso3, target = Q2.5_iso3) |>
  unnest(target) |> 
  mutate(across(c(home, target), 
                list(continent = ~countrycode(., "iso3c", "continent",
                                              custom_match = c("XKX" = "Europe"))))) |> 
  mutate(across(ends_with("_continent"),
                ~recode(., Oceania = "Asia & Oceania", Asia = "Asia & Oceania"))) |> 
  group_by(home_continent, target_continent) |>
  summarize(times_marked = n()) |> 
  ungroup() |> 
  mutate(id = 1:n()) |> 
  pivot_longer(ends_with("continent"), names_to = "x", values_to = "continent") |> 
  mutate(continent_total = sum(times_marked), .by = x) |> 
  group_by(x, continent) |> 
  mutate(prop = sum(times_marked) / continent_total) |> 
  ungroup() |> 
  mutate(nice_label = glue(
    '{continent}<br><span style="font-size: 6.5px">{label_percent(accuracy = 1)(prop)}</span>'
  ))

ggplot(
  continents_marked,
  aes(x = x, y = times_marked, stratum = continent, alluvium = id, fill = continent)
) +
  geom_flow() +
  geom_stratum(color = "grey30", alpha = 0.5, linewidth = 0.1) +
  geom_richtext(
    fill = NA, label.color = NA, label.padding = grid::unit(rep(0, 4), "pt"),
    stat = "stratum", aes(label = nice_label, color = continent), 
    size = 2.2, lineheight = 0.9,
    family = "Fira Sans Semibold"
  ) +
  scale_x_discrete(
    labels = c("Home region", "Target region"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(
    values = c(clrs$Prism[6], clrs$Prism[8], clrs$Prism[11], clrs$Prism[2]),
    guide = "none"
  ) +
  scale_color_manual(
    values = colorspace::darken(c(
      clrs$Prism[6], clrs$Prism[8],
      clrs$Prism[11], clrs$Prism[2]
    ), 0.5),
    guide = "none"
  ) +
  labs(x = NULL, y = NULL) +
  theme_ingo() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.x = element_text(family = "Fira Sans Semibold", size = rel(0.7)),
    axis.text.y = element_blank()
  )
```

While comparatively few INGOs are headquartered outside the Global North, these organizations have a worldwide reach. In total, INGOs answered questions about their work in 148 different countries. The regional distribution of INGO target countries reflects their broader scope and is almost a mirror image of these organizations' home regions—nearly two-thirds of respondents answered questions about their work in Africa and Asia (see right panel of @fig-work-hq-regions), the most underrepresented regions for NGO headquarters. This disparity could reflect the fact that INGOs from the Global South without a clear English-language internet presence were excluded from the sample, but the pattern also corresponds to similar findings in other surveys of global NGOs [@Limoncelli:2016; @HeissKelley:2016; @FootToftCesare:2015], reinforcing the idea that Northern-based NGOs tend to focus their work in the South. Organizations also operate in a variety of political contexts. Unsurprisingly, given the concentration of INGOs in the Global North, the vast majority of respondents (90%) are headquartered in democracies. The countries these organizations work in, though, show more diversity in regime type—a third of respondents answered questions about their work in autocratic countries.

@@

Building on these critiques, a new trend emerged that treated transnational actors as simultaneously normative and strategic [@Cooley:2002; @Keck:1998; @Kelley:2009a]. @MitchellSchmitz:2014 summarized this debate with their theory of nonprofit principled instrumentalism, or the notion that INGOs must balance their normative mission-driven principles against the instrumental needs of organizational survival. Unlike governments that work toward producing social value (such as maximizing national security or promoting general welfare) and unlike firms that focus on financial performance (such as maximizing profits for shareholders), nonprofit organizations face a tradeoff between their social mission and their need for resources and must pursue both concurrently—they must "instrumentally pursue their principled objectives within the economic constraints and political opportunity structures imposed by their external environments" [@MitchellSchmitz:2014, 489].

INGOs are motivated primarily by their foundational mission and values, and they structure their strategies around this vision to ensure that their programming furthers their mission [@Moore:2000]. Survey respondents confirm this motivation, describing their core mission both in aspirational terms, including "fighting poverty and holding leaders accountable,"[^1193] encouraging "awareness and pro-action towards sustainable development,"[^1267] and promoting "democracy and open society" and "governments' respect for the human rights of children";[^1328_1252] and through more specific statements, such as "improv[ing] healthcare, specifically neurological services, in the least developed nations,"[^1257] or ending "the use of immigration detention around the world."[^1376]

Most respondents report a tight link between programs and missions, stating that "[o]ur mission, values and vision are reflected in everything that we do," that "all our work is geared towards the accomplishment of this objective," and that "our approach matches our mission precisely."[^1245_1267_1272] For many organizations, the "mission and values are the reason we work on these objectives"[^1643]—the normative goals of the organization provide strong motivation to engage in pro-social advocacy and services. Specific programming strategies are also closely linked to INGO missions. For instance, an organization working in South Asia has ambitions to "[b]uild the capacity of local organisations to equip adolescent girls to access their rights and address gender inequality through sport." As expected, much of their programming is centered on sports education and leadership training, but importantly, the organization makes a point that nothing they do is implemented directly: because their core mission is to build local capacity, they ensure that "all the work we do is through local organisations."[^1799] Organizations thus work to ensure that their strategies align with and reflect their core missions, visions, and values.

@@

The distance between the government's preferences for a given issue and an INGO's core mission and vision for that issue determine how contentious their relationship will be, ranging from cooperative to antagonistic. This distance, in turn, influences how governments regulate INGO work. The alignment of organizational missions and strategies with government preferences, therefore, is crucial for understanding government regulation of INGOs. Organizations working on issues that align with regime preferences will be treated more favorably and with less regulatory pressure than INGOs working on more threatening, less aligned issues.

@@

According to @MitchellSchmitz:2014, nonprofit organizations are primarily *driven* by their missions and *constrained* by instrumental concerns: worries about money and other resources get in the way of pursuing deeper normative principles. I argue, however, that this view of instrumentality is insufficient. Instrumental concerns, including time, money, or political connections, are not simply constraints. These resources are what allow INGOs to pursue their principled ideals in the first place—they provide organizations with the flexibility to engage in programming and adapt to the political and regulatory environments they face. The region surrounding the organization's ideal point represents its reserve of programmatic flexibility, or the organization's ability and willingness to make changes to its programs in response to regulations.

@@

Given their frequency, the regulatory changes INGOs observe do not necessarily correlate with formal statutory developments. These laws give bureaucrats and politicians significant interpretive leeway, however, and allow them to reshape the regulatory environment for international NGOs [@ChaudhryHeiss:2022]. Because of this dynamic discretionary de facto environment, INGOs actively seek out information about changes in laws, rules, and regulations. International NGOs are most likely to hear about new legal changes from other NGOs in the country, both from their local partners and from other foreign organizations. They also receive news about changes directly from government officials, sometimes at inopportune moments. One INGO working in East Africa, unaware of any changes in regulations, discovered new funding requirements for foreign NGOs when requesting permits to expand their operations in the country—the organization ultimately decided to not expand as a result.[^52] INGOs also find out about changes in laws through the media and through internet research, including websites that aggregate information about civil society regulations like Civicus and the International Center for Not-for-Profit Law (ICNL).

@@

---

Can apply to other situations, like looking at regime laws or other stuff like that

@@

[^1193]: Response 1193.

[^1267]: Response 1267.

[^1328_1252]: Responses 1328 and 1252.

[^1257]: Response 1257.

[^1376]: Response 1376.

[^1245_1267_1272]: Response 1245, 1267, and 1272.

[^1643]: Response 1643.

[^1799]: Response 1799.

@@

\newpage

# References

::: {#refs}
:::

\newpage

:::: {#appendix-count}

# Appendix

::::
